---
title: "CCE 102 Waitlist"
author: "Greg Macfarlane"
date: "2025-07-07"
categories: [Teaching]
---

```{r libs, include = FALSE}
library(tidyverse)
library(kableExtra)
library(wesanderson)
library(readxl)
pal <- wes_palette("AsteroidCity1")
theme_set(theme_bw())

clean_major <- function(x) {
  x <- trimws(x)
  case_when(
    x == "Civil Engineering" ~ "Civil Engineering",
    x == "Construction Management" ~ "Construction Management",
    x == "Facility and Property Management" ~ "Facilities Management",
    str_detect(x, "Engineering") ~ "Other Engineering",
    TRUE ~ "Other"
  )
}

clean_class <- function(x) {
  x <- gsub("\u00a0", "", x)
  x <- trimws(x)
  factor(trimws(x), levels = c("Senior", "Junior", "Sophomore", "Freshman"))
}
```

# Background
Beginning with the 2024-2025 academic year, CCE 102 replaces a two different 200-level classes. This creates 
a logjam where two years of students need to be enrolled at the same time. CFM, CE, and FM graduate together about 
150 students per year, meaning that sets the long-run capacity for the class. Our plan is to offer four sections 
of 60 students per year, with a total capacity of 240 students. This should provide ample space for students who 
take the class and then switch majors, as well as provide a context for students in other majors who take the class 
as a GE or a minor elective. 

During the 2024-2025 academic year, we offered six sections of CCE 102 --- in addition to a Spring study abroad 
section --- with a total capacity of 380 students. This should have been sufficient to clear the backlog of students for 
two class years. However, the waitlist for the 2025-2026 academic year remains large. 

This post serves as an analysis of the waitlist for Fall 2025, with an eye to understanding who 
is yet to enroll in the course.

```{r folder, include = FALSE}
folder <- c("/Users/gregmacfarlane/Library/CloudStorage/Box-Box/Macfarlane/2 Areas/cce102/waitlist-summer25")
files <- dir(folder)
```

## An estimate of the backlog
I asked Emily Platt to run a report with the university registrar, returning a list of students who are declared
CE, CFM, or FM majors but who have not yet completed CCE 102 or its predecessor course CCE 201. 

```{r backlog}
backlog <- read_xlsx(file.path(folder, "selected_unenrolled.xlsx")) |>
  mutate(
    # simplify major categories
    major = clean_major(Major),
    # mark new students
    new = ifelse(`BYU Total Credits` < 12, TRUE, FALSE),
  )
```

The total number of students who are in the majors but who have not yet enrolled is `r nrow(backlog)`.
@tbl-backlog shows the number of students who have not yet enrolled in CCE 102 by major.
About 1/3 of the backlog  (`r nrow(backlog |> filter(new))` students) are recently enrolled students, meaning that they have completed fewer than
12 BYU credit hours. These students could be new to BYU as freshmen, or they
could have transferred from another school. 
 

```{r}
#| label: tbl-backlog
#| tbl-cap: Students without CCE 102
backlog |>
  group_by(major, new) |>
  summarise(n = n()) |>
  pivot_wider(names_from = new, values_from = n, values_fill = 0) |>
  kbl(col.names = c("Major", "More than 12 credit hours", "Less than 12 credit hours")) |> 
  kable_styling()
```

@fig-backlog shows the distribution of students who have not yet enrolled in CCE 102 by number of credits,
excluding students with fewer than 12 credit hours. The most common number of hours is less than 50,  but there are some older
students particularly in CM.

```{r }
#| label: fig-backlog
#| fig-cap: Distribution of not new students who have not yet enrolled in CCE 102 by number of credits.
ggplot(backlog |> filter(!new), aes(x = `BYU Total Credits`, fill = major)) +
  geom_histogram(binwidth = 12)  + scale_fill_manual("Major", values = pal[c(1, 3, 4)]) 
```

## Current enrollment trends
Of course, the question remains about how many of these students are currently registered for the course.
I obtained the waitlist and enrollment information for my own classes, and asked Darrell Sonntag to send me his.

```{r enrollment, include = FALSE}
mac_enrollment <- read_csv(file.path(folder, "macfarlane_roll.csv")) |>
  transmute(id = str_remove_all(id, "-"), major = clean_major(major),
            instructor = "Macfarlane", status = "enrolled", class = clean_class(class))
mac_waitlist <- read_delim(file.path(folder, "macfarlane_waitlist.tsv"), delim = "\t") |>
  transmute(id = str_remove_all(`BYU ID`, "[:space:]|-"), major = clean_major(`Major/Emphasis`),
            instructor = "Macfarlane", status = "waitlist", class = clean_class(`Standing`))
  
sonntag_enrollment <- read_xlsx(file.path(folder, "sonntag_roll.xlsx")) |>
  transmute(id = str_remove_all(`BYU ID`, "-"), major = clean_major(Emphasis),
            instructor = "Sonntag", status = "enrolled", class = clean_class(Standing))
sonntag_waitlist <- read_xlsx(file.path(folder, "sonntag_waitlist.xlsx")) |>
  transmute(id = str_remove_all(`BYU ID`, "[:space:]|-"), major = clean_major(`Major/Emphasis`),
            instructor = "Sonntag", status = "waitlist", class = clean_class(Standing))

enrollment <- bind_rows(mac_enrollment, mac_waitlist, sonntag_enrollment, sonntag_waitlist) 
```

First off, let's get a list of unique students, because some students may be waitlisted for one section while enrolled
in another. Looks like there are 5 students who are on the waitlist for Macfarlane who are enrolled in Sonntag's class.

```{r}
# list of students who appear more than once
double_e <- enrollment |> group_by(id) |> mutate(n = row_number()) |> filter(n > 1) |> pull(id)
enrollment |> filter(id %in% double_e) |> arrange(id)
```

```{r }
#| label: tbl-unique
#| tbl-cap: Students Enrolled and Waitlisted for Fall 2025
unique <- enrollment |> group_by(id) |> summarise(n = n()) |> filter(n == 1) |> pull(id)
u_enrollment <- enrollment |> filter(id %in% unique)

u_enrollment |> group_by(major, class, status) |>
  summarise(n = n()) |>
  pivot_wider(names_from = major, values_from = n, values_fill = 0)  |>
  arrange(status, class) |>
  select(-status) |>
  kbl(col.names = c("Class", "CE", "CM", "FM", "Other", "Other Engineering"))  |>
  kable_styling() |>
  group_rows("Enrolled", 1, 4) |>
  group_rows("Waitlisted", 5, 8)
```

This means that there are currently `r nrow(u_enrollment)` unique students enrolled and waitlisted in CCE 102, with both
sections at the capacity of 60 students each.

## Overlap with Backlog
It remains to be seen how many of the students on the enrollment list are also on the backlog.

```{r }
#| label: tbl-remaining
#| tbl-cap: Remaining Backlog by Major
remaining_backlog  <- backlog |>
  left_join(u_enrollment |> select(id, status), by = c(`BYU ID` = "id")) |>
  mutate(status = ifelse(is.na(status), "not enrolled", status))

remaining_backlog |>
  group_by(major, status) |>
  summarise(n = n())  |>
  pivot_wider(names_from = status, values_from = n, values_fill = 0)  |>
  kbl(col.names = c("Major", "Not Enrolled", "Waitlisted"))  |>
  kable_styling()
```

Ugh. It seems like only students who are not enrolled in CCE 102 are on the backlog. Let's look again at our distribution
of credit hours by backlog status, which I show in @fig-remaining. I can't show the hour distribution for the enrolled 
students, because the information on number of credit hours came from Emily Platt's report and not the enrollment data.

```{r }
#| label: fig-remaining
#| fig-cap: Distribution of not new students who have not yet enrolled in CCE 102 by number of credits.
ggplot(remaining_backlog |> filter(!new), aes(x = `BYU Total Credits`, fill = major)) +
  geom_histogram(binwidth = 12)  + scale_fill_manual("Major", values = pal[c(1, 3, 4)])  + 
  facet_wrap(~ status)
```

## Recommendation
This is the first time we have ever run a report like this, so we don't know if this
is how all classes look, or if it just this one. Some of the students on the backlog may be 
waiting to declare a different major or withdraw from the university, and therefore show
up on report but are not actively engaged with the major. It is confusing to me that
so few of the incoming new students are enrolled or waitlisted, and also that
there are so many students on the backlog including those who have to have been
here for years.

My recommendation is that we modestly increase the capacity of the class for Fall 2025. 
I can accommodate 72 students in my section, because I can increase the size of each team.
I cannot increase the number of teams, because all teams have to make a presentation in 
class.

If the overenrollment continues in Winter 2025, we may have to add an additional section. 
But the added capacity from 2024-2025 should have been sufficient to clear the backlog 
of students who are actively engaged in the major.
