---
title: "CCE 102 Grade Analysis"
author: "Greg Macfarlane"
date: "2025-04-29"
categories: [Teaching]
---

```{r libs, include = FALSE}
library(tidyverse)
library(kableExtra)
library(wesanderson)
pal <- wes_palette("AsteroidCity1")
```
```{r load, include = FALSE}
folder <- c("/Users/gregmacfarlane/Library/CloudStorage/Box-Box/Macfarlane/2 Areas/cce102/grades")
files <- dir(folder)
# read files that contain enrollment information
enrollment <- lapply(files[str_detect(files, "enrollment")], function(x) {
    read_csv(file.path(folder, x))  |>
    set_names(c("ln", "fn", "netid", "byuid", "section", "major")) |>
    select(netid, section, major)
}) |>
  set_names(c("Fall 2024", "Winter 2025")) |>
  bind_rows(.id = "semester") 

# read files that contain grade information
grades <- lapply(files[str_detect(files, "grades")], function(x) {
    read_csv(file.path(folder, x)) |>
    set_names(c("netid", "ln", "fn", "current_pct", "current_grd", "final_grd", "final_gpa")) |>
    select(netid, current_pct, current_grd, final_grd, final_gpa) |>
    mutate(current_pct = as.numeric(str_remove(current_pct, "%")))
}) |>
  bind_rows()  

data <- enrollment |>
  left_join(grades, by = "netid")  |>
  mutate(major = case_when(
    major == "Civil Engineering" ~ "Civil Engineering",
    major == "Construction and Facilities Management" ~ "Construction and Facilities Management",
    TRUE ~ "Other"
  ))
```

I taught four sections of CCE 102 this year. This post serves to document and
record how the students performed in each section and by major. Across the four sections,
`r nrow(data)` students completed the course. The average GPA by section has not been 
consistent across all four sections, however.

```{r enrollment}
data |>
  group_by(semester, section) |>
  summarise(count = n(), `gpa` = mean(final_gpa)) |>
  kbl(digits = 2, col.names = (c("Semester", "Section", "Count", "Average GPA"))) |> 
  kable_styling()
```

As a note, I am only including the courses where I was the primary instructor.
Darrell Sonntag taught an additional section each semester, but he and I had
slightly different assignments and grading policies to the point where I don't think
it would be productive to compare them.

Let's look at the distribution of grades by section (see @fig-gradedist). These
percentile grades do not include any final curve adjustments. Each section
reached its GPA in different ways. Fall 2024  section 003 had a number of
students earning low but passing C grades, while Winter 2025 001 had a few
students perform poorly and earn E grades.

```{r distr}
#| label: fig-gradedist
#| fig-cap: Grade distribution by section.
ggplot(data, aes(x = current_pct, color = str_c(semester, section, sep = " "))) + 
  geom_density(alpha = 0.25) +
  scale_color_manual("Section", values = pal) +
  theme_bw() + xlab("Un-curved Final Grade") + ylab("Density")
```

Winter 2025 felt much better in terms of student attitude and workload. At the same
time, the average grades were a little bit lower. I'm curious how the student evaluations
shake out.

Now let's consider the breakdown across majors, which I show in @fig-grademaj. Note that
this is a GE class, with `r nrow(table(enrollment$major))` different majors enrolled. However,
92% of students are enrolled in either Civil Engineering (CE) or in Construction
and Facilities Management (CFM), where this class is required. CE
students do better in the class, with the mode being about an A-. CFM majors have
a wide distribution, but the mode is in the B range. Other majors --- including
a few environmental science students and multiple open majors --- perform more
like CE students, with a few poor performing exceptions.

```{r distr}
#| label: fig-grademaj
#| fig-cap: Grade distribution by section.
#| 
ggplot(data, aes(x = current_pct, color = major)) + 
  geom_density(alpha = 0.25) +
  scale_color_manual("Major", values = pal) +
  theme_bw() + xlab("Un-curved Final Grade") + ylab("Density")
```


The final grades awarded by all semesters (including with a curve applied) are 
shown in @fig-final. The most common grade awarded to all students in all majors is an A,
particularly when combined with the A- grades. But there are probably too many CFM
students earning B- and C grades to make me happy. 

```{r final}
#| label: fig-final
#| fig-cap: "Final Grades awarded by major, all semesters."
data |>
    mutate(final_grd = factor(final_grd, levels = c("A+", "A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "D-", "E", "W"))) |>
ggplot(aes(x = final_grd, fill = major)) +
    geom_bar(, position = "dodge") + 
    xlab("Final Grade") + ylab("Number of Students")
    theme_bw()
```

The chart above does not contain withdrawals, as students who withdraw do not remain on my rolls. I 
cannot therefore track what major the withdrawn students are enrolled in. But I can compute the
overall DEW rate for both semesters, shown in @tbl-dew. This rate has come down, but 
it would be nice to get it below 5%.

```{r dew}
#| label: tbl-dew
#| tbl-cap: DEW Rate by Semester
# withdrawals
withdrawals <- tibble(semester = c("Winter 2025", "Fall 2024"), w = c(3, 4))

data |>
    mutate(de = case_when(
      str_detect(final_grd, "D") ~ TRUE,
      str_detect(final_grd, "E") ~ TRUE,
      TRUE ~ FALSE
    )) |>
    group_by(semester) |>
    summarise(n = n(), de = sum(de))  |>
    left_join(withdrawals, by = "semester") |>
    mutate(dew = (de + w) / (n + w) * 100)  |>
    kbl(col.names = c("Semester", "N Complete", "D/E", "W", "DEW [%]"), digits = 2)  |>
    kable_styling()
```
